---
title: "Canonical Correlation Analysis"
output: html_notebook
---

Data: Next generation sequencing serum miRNA content and nightingale-metabolomics



Setup working directory and install mixOmics package

```{r}

## install BiocManager if not installed if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager") 
## install mixOmics 
BiocManager::install('mixOmics')
```

Step 1 Load  and edit data
```{r}

# Import data 
cmiR1 <- read.delim("normalized_miR_counts.txt")
transposed_counts <- t(cmiR1) #samples need to be as rows
cmiR <- as.data.frame(transposed_counts)
mets <- read.table("boxcox_transf_data.txt")
metadata <-read.delim("phenoData_filtered.txt")


```

PLS
```{r}
library(mixOmics) # import the mixOmics library

X <- cmiR # use the cmiR expression data as the X matrix
data.matrix(X, rownames.force = NA)
Yy <- mets # use the metabolic data as the Y matrix
data.matrix(Yy, rownames.force = NA)

## Check NA values from matrix
sum(is.na(X)) # number of cells with NA
sum(is.na(Yy)) # number of cells with NA

#if needed input missing values, NIPALS is used to decompose the dataset. 
Y <- impute.nipals(X = Yy, ncomp = 10)
sum(is.na(Y)) # number of cells with NA

pls.result <- pls(X, Y) # run the method
plotIndiv(pls.result)   # plot the samples
plotVar(pls.result)     # plot the variables

```

Preliminary Analysis with PCA
```{r}
library(mixOmics)
pca.mir <- pca(X, ncomp = 10, center = TRUE, scale = TRUE)
pca.metabolomics <- pca(Y, ncomp = 10, center = TRUE, scale = TRUE)

plot(pca.mir)
plot(pca.metabolomics)
```
```{r}
library(mixOmics)
plotIndiv(pca.mir, comp = c(1, 2), 
          group = ordered.phenodata$Cancer_during_surveillance, 
          ind.names = ordered.phenodata$Cancer_during_surveillance, 
          legend = TRUE, title = 'miR, PCA comp 1 - 2')

plotIndiv(pca.metabolomics, comp = c(1, 2), 
          group = ordered.phenodata$Cancer_during_surveillance, 
          ind.names = ordered.phenodata$Cancer_during_surveillance, 
          legend = TRUE, title = 'Metabolomics, PCA comp 1 - 2')
```


sPLS tuning
```{r}
library(mixOmics)
#initial sPLS model
spls.LS <- spls(X = X, Y = Y, ncomp = 5, mode = 'regression')

#Tuning sPLS
# repeated CV tuning of component count
perf.spls.LS <- perf(spls.LS, validation = 'Mfold',
                         folds = 10, nrepeat = 5) 

plot(perf.spls.LS, criterion = 'Q2.total')

# set range of test values for number of variables to use from X dataframe
list.keepX <- c(seq(10, 50, 5))
# set range of test values for number of variables to use from Y dataframe
list.keepY <- c(3:10) 


tune.spls.LS <- tune.spls(X, Y, ncomp = 2,
                              test.keepX = list.keepX,
                              test.keepY = list.keepY,
                              nrepeat = 1, folds = 10, # use 10 folds
                              mode = 'regression', measure = 'cor') 
plot(tune.spls.LS)         # use the correlation measure for tuning

tune.spls.LS$choice.keepX

tune.spls.LS$choice.keepY

# extract optimal number of variables for X dataframe
optimal.keepX <- tune.spls.LS$choice.keepX 

# extract optimal number of variables for Y datafram
optimal.keepY <- tune.spls.LS$choice.keepY

optimal.ncomp <-  length(optimal.keepX) # extract optimal number of components

```
Final model, Using the tuned parameters generated above, the final sPLS model can be constructed.
```{r}
# use all tuned values from above
final.spls.LS <- spls(X, Y, ncomp = optimal.ncomp, 
                    keepX = optimal.keepX,
                    keepY = optimal.keepY,
                    mode = "regression") # explanitory approach being used, 
                                         # hence use regression mode

```

Plots1, Sample plot for sPLS2 performed on the LS data. Samples are projected into the space spanned by the components associated to each data set (or block).
```{r}
plotIndiv(final.spls.LS, ind.names = FALSE, 
         rep.space = "X-variate", # plot in X-variate subspace
         group = ordered.phenodata$Sex, # colour bytype
         pch = as.factor(ordered.phenodata$Cancer_during_surveillance),  #  By status
         col.per.group = color.mixo(1:2), 
         legend = TRUE, legend.title = 'Sex', legend.title.pch = 'Type')

plotIndiv(final.spls.LS, ind.names = FALSE,
         rep.space = "Y-variate", # plot in Y-variate subspace
         group = ordered.phenodata$Sex, # colour by sex
         pch = as.factor(ordered.phenodata$Cancer_during_surveillance), 
         col.per.group = color.mixo(1:2), 
         legend = TRUE, legend.title = 'Type', legend.title.pch = 'Health')
```



Plots 2.Sample plot for sPLS2 performed on the LS data. Samples are projected into the space spanned by the averaged components of both datasets (cmiR and metabolomics).
```{r}
plotIndiv(final.spls.LS, ind.names = FALSE, 
         rep.space = "XY-variate", # plot in averaged subspace
         group = ordered.phenodata$Sex, # colour by Type
         pch = as.factor(ordered.phenodata$Cancer_during_surveillance), # symbol per health status
         col.per.group = color.mixo(1:2),                      # by dose group
         legend = TRUE, legend.title = 'Sex', legend.title.pch = 'Status')

```


Plot 3, The start of the arrow indicates the location of a given sample in the space spanned by the components associated to the miR data set, and the tip of the arrow the location of that same sample in the space spanned by the components associated to the metabolism data set
```{r}
plotArrow(final.spls.LS, ind.names = FALSE,
          group = ordered.phenodata$Cancer_during_surveillance, # colour by status
          col.per.group = color.mixo(1:8),
          legend.title = 'Sex')
```


Variable plots 1, The stability of a given feature is defined as the proportion of cross validation folds (across repeats) where it was selected for to be used for a given component.
```{r}
# form new perf() object which utilises the final model
perf.spls.LS <- perf(final.spls.LS, 
                          folds = 5, nrepeat = 10, # use repeated cross-validation
                          validation = "Mfold", 
                          dist = "max.dist",  # use max.dist measure
                          progressBar = FALSE)

# plot the stability of each feature for the first two components, 
# 'h' type refers to histogram
par(mfrow=c(1,2)) 
plot(perf.spls.LS$features$stability.X[[1]], type = 'h',
     ylab = 'Stability',
     xlab = 'Features',
     main = '(a) Comp 1', las =2,
     xlim = c(0, 150))
plot(perf.spls.LS$features$stability.X$comp2, type = 'h',
     ylab = 'Stability',
     xlab = 'Features',
     main = '(b) Comp 2', las =2,
     xlim = c(0, 300))
```

Variable plots 2, The relationship between the features and components can be explored using a correlation circle plot. This highlights the contributing variables that together explain the covariance between the two datasets. Specific subsets of molecules can be further investigated. This plot shows the correlations between selected miRs , between selected metabolic parameters and the relationship between sets of miRs and certain metabolic parameters.
```{r}
plotVar(final.spls.LS, cex = c(3,4),cutoff = 0.45, var.names = c(TRUE, TRUE))
```

variable plots 3, The networks are bipartite, where each edge links a mir (rectangle) to a metabolomic variable (circle) node, according to a similarity matrix.Saves image
```{r}
 # Load necessary package
library(mixOmics)

# Set the colors for the connecting lines
color.edge <- color.GreenRed(50)

# Open a new window with specified size (width and height in inches)
X11(width = 10, height = 8)

# Define the path and name of the saved PNG file
png(filename = "sPLS_Lynch_Study_Network_Plot_LS_H.png", width = 2000, height = 1600, res = 200)

# Generate the network plot
network(final.spls.LS, comp = 1:2,
        cutoff = 0.30, # only show connections with a correlation above 0.30
        shape.node = c("rectangle", "circle"),
        color.node = c("orange", "lightblue"),
        color.edge = color.edge)

# Close the PNG device
dev.off()

# Close the X11 device
dev.off()

#color.edge <- color.GreenRed(50)  # set the colours of the connecting lines

#X11() # To open a new window for Rstudio
#network(final.spls.LS, comp = 1:2,
 #       cutoff = 0.30, # only show connections with a correlation above 0.30
  #      shape.node = c("rectangle", "circle"),
   #     color.node = c("orange", "lightblue"),
    #    color.edge = color.edge,
     #   save = 'png', # save as a png to the current working directory
      #  name.save = 'sPLS Lynch Study Network Plot_LS_H')
```
Variable plots 4, The heatmap plot displays the similarity values between the cmir and metabolomics variables selected across two dimensions, and clustered with a complete Euclidean distance method. Saves image
```{r}
# Load necessary package
library(mixOmics)

# Open a new window with larger size (width and height in inches)
X11(width = 20, height = 15)

# Define the path and name of the saved JPEG file with larger dimensions
jpeg(filename = "PLS_CIM_image_LS_H_0.4.jpeg", width = 4000, height = 3000, res = 300)

# Set graphical parameters for text size
par(cex.main = 1, cex.lab = 1, cex.axis = 1, mar = c(15, 15, 15, 15))

# Generate the heatmap plot with adjusted text sizes
cim.res <- cim(final.spls.LS, comp = 1:2, cutoff = 0.3,
               xlab = "Metabolism", ylab = "cmiRs",
               save = "jpeg", name.save = "PLS_CIM_image_LS_H_0.4")


```


